Tree biodiversity and ecosystem function relationship across southern African woodlands

### 2020-09-14 manuscript/regional_befr_manuscript.tex
Added minor comments to manuscript

% Savannas might not be the same as forests
Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. A threshold stem density may exist below which the effects of biodiversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.

@@

% African savannas are understudied in this regard
The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants. A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.

% \citep{Michaletz2014, Michaletz2018} \citep{Spasojevic2014}
%In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
% Due the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass. 
% Environment affects the BEF relationship, so must be accounted for.
Environmental variation is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}. Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}. It is important therefore to account for environment and understand how it interacts with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas. \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.

@@

Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\fmbetadhb{}) when environmental covariates were accounted for. Again, the direct effect of species diversity on AGB was negligible (\fmbetadb{}). The total effect of species diversity on AGB was positive (\fmbetatotaldb{}). Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased (), but the predictive power of the model as a whole increased ().

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density, however, the direct effect on AGB remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales \citep{Muledi2017, Bucini2007} in southern African woodlands, often being further structured by the vegetation overlying it, an aspect which SoilGrids does not model precisely. Due to the plots used in this study often being situated non-randomly in the landscape, coupled with the coarseness of the SoilGrids data, it is not surprising that this model path is poorly constrained. Soil data is time-consuming to collect and difficult to compare across studies when different protocols are used, though this study prompts the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands. \citet{Lehmann2014} similarly found weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, fire and tree basal area.

>>>>

% Savannas might not be the same as forests
Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship. Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. A threshold stem density may exist below which the effects of tree species diversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.

@@

% African savannas are understudied in this regard
The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants. A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>100s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.

% \citep{Michaletz2014, Michaletz2018} \citep{Spasojevic2014}
%In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
% Due the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass. 
% Environment affects the BEF relationship, so must be accounted for.
Environmental heterogeneity is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}. Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}. It is important therefore to account for this environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas. \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.

@@

Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\fmbetadhb{}) when environmental covariates were accounted for. Again, the direct effect of species diversity on AGB was negligible (\fmbetadb{}). The total effect of species diversity on AGB was positive (\fmbetatotaldb{}). Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased, but the predictive power of the model as a whole increased.

@@

A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass. The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density. The direct effect on AGB however, remained positive and marginally significant, as expected. Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables. This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used. SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales \citep{Muledi2017, Bucini2007} in southern African woodlands, often being further structured by the vegetation overlying it, an aspect which SoilGrids does not model precisely. Due to the plots used in this study often being situated non-randomly in the landscape, coupled with the coarseness of the SoilGrids data, it is not surprising that this model path is poorly constrained. Soil data is time-consuming to collect and difficult to compare across studies when different protocols are used, though this study prompts the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands. \citet{Lehmann2014} similarly found weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, fire and tree basal area.
